Demo: Counterpoint Analysis and Synthesis

### 2022-08-16 doc/farm22/abstract.tex
figures complete

For simplicity we now focus only on the top two voices. Feeding
Beethoven's notes into the tool as two part first species
counterpoint, it easily finds and reports errors in the use of perfect
fourths, missing octaves at the boundaries, and similar motion into
fifths and octaves. We can disable the boundary rules as they are not
relevant and relax the consonance rule to allow perfect fourths, leaving
only the motion errors. We can replace the note F that Haydn fixed
with a hole, and run the tool in synthesis mode using the same
rules. It makes the same fix Haydn did. Perhaps we gave it too much
guidance, so instead we can create three holes in a row near the
error, which generates the solution shown in
Figure~\ref{fig:b146fix3}, with the three generated notes inside the
blue box. Although there are several possible solutions, the SMT
solver has chosen to retain the third A while changing the first two
notes to A as well; the only difference with Haydn's correction is
that the first note was changed from G to A as well. Although this
creates a satisfying oblique motion into a perfect fifth between bars
2 and 3, this same motion occurs between bars 5 and 6; furthermore we
now have four A notes in a row in addition to the three G notes at the
end of the second voice. For a middle voice this is not too bad, but
viewed as just two voices it is a big dull.

\begin{figure}
  \includegraphics[width=8cm]{figures/b146fix3.png}
  \caption{Three Notes Changed}
  \Description{Three Notes Changed}
  \label{fig:b146fix3}
\end{figure}

This can be fixed by adding a constraint to limit the number of repeated
notes, but we can also try generating our own two part counterpoint
following Haydn's cantus firmus, fixing only the first and last of
Beethoven's notes. To help create better quality counterpoint we add
constraints to limit the number of leaps (horizontal intervals of more
than a major third) to at most one and require at least six instances
of contrary motion, which turns out to be maximal. Note that it would
be extremely difficult for a human being with no computer assistance
to generate counterpoint following such restrictions, but the SMT
solver instantly returns the solution shown in
Figure~\ref{fig:b146gen3}.

\begin{figure}
  \includegraphics[width=8cm]{figures/b146gen3.png}
  \caption{Generated Counterpoint}
  \Description{Generated Counterpoint}
  \label{fig:b146gen3}
\end{figure}

>>>>

For simplicity we now focus only on the top two voices. Feeding
Beethoven's notes into the tool as two part first species
counterpoint, it easily finds and reports errors in the use of perfect
fourths, missing octaves at the boundaries, and similar motion into
fifths and octaves. We can disable the boundary rules as they are not
relevant and relax the consonance rule to allow perfect fourths, leaving
only the motion errors. We can replace the note F that Haydn fixed
with a hole, and run the tool in synthesis mode using the same
rules. It makes the same fix Haydn did. Perhaps we gave it too much
guidance, so instead we can create three holes in a row near the
error, which generates the solution shown in
Figure~\ref{fig:b146fix3}, with the three synthesized notes inside the
blue box. The only difference with Haydn's correction is
that the first note was changed from G to A as well. Although this
creates a satisfying oblique motion into a perfect fifth between bars
2 and 3, this same motion occurs between bars 5 and 6; furthermore we
now have four A notes in a row in addition to the three G notes at the
end of the second voice. For a middle voice this is not too bad, but
viewed as just two voices it is a bit dull.

\begin{figure}
  \includegraphics[width=8cm]{figures/b146fix3.png}
  \caption{Three Notes Synthesized}
  \Description{Three Notes Synthesized}
  \label{fig:b146fix3}
\end{figure}

This can be fixed by adding a constraint to limit the number of repeated
notes, but we can also try generating our own two part counterpoint
following Haydn's cantus firmus, fixing only the first and last of
Beethoven's notes. To help create better quality counterpoint we add
constraints to limit the number of leaps (horizontal intervals of more
than a major third) to at most one and require at least six instances
of contrary motion, which turns out to be maximal. We also disallow
perfect fourths except at the boundary. Note that it would
be extremely difficult for a human being with no computer assistance
to generate counterpoint following such restrictions, but the SMT
solver instantly returns the solution shown in
Figure~\ref{fig:b146gen3}. Contrary motion is shown by pairs of blue
arrows and the single leap by a red line.

\begin{figure}
  \includegraphics[width=8cm]{figures/b146gen.png}
  \caption{Generated Counterpoint}
  \Description{Generated Counterpoint}
  \label{fig:b146gen3}
\end{figure}
